\documentclass[preprint]{sigplanconf}

\usepackage{inconsolata}
\usepackage{hyperref}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{mathtools}
\usetikzlibrary{chains,fit,shapes,calc}
\usepackage{semantic}
\usepackage{tabu}
\usepackage{amsthm}
\usepackage{mathptmx}
\usepackage{todonotes}
\usepackage{listings}
\lstset{
  frame=none,
  xleftmargin=6pt,
  stepnumber=1,
  belowcaptionskip=\bigskipamount,
  captionpos=b,
  escapeinside={*'}{'*},
  language=haskell,
  tabsize=2,
  emphstyle={\bf},
  commentstyle=\it\ttfamily,
  stringstyle=\mdseries\ttfamily\color{violet},
  showspaces=false,
  keywordstyle=\bfseries\ttfamily\color{teal},
  columns=flexible,
  basicstyle=\small\ttfamily,
  showstringspaces=false,
  morecomment=[l]\%,
}
\lstnewenvironment{code}{}{}

\begin{document}
\title{Shared Environment Call-By-Need}
\subtitle{A Functional Pearl}

\authorinfo{George Stelle}
           {University of New Mexico}
           {stelleg@cs.unm.edu}
\authorinfo{Darko Stefanovic}
           {University of New Mexico}
           {darko@cs.unm.edu}
\authorinfo{Stephen L. Olivier}
           {Sandia National Laboratories}
           {slolivi@sandia.gov}
\authorinfo{Stephanie Forrest}
           {University of New Mexico}
           {forrest@cs.unm.edu}

\maketitle

\begin{abstract}
  Existing machines for lazy evaluation use a \emph{flat}
  representation of environments, storing the terms associated with
  free variables in an array.  Combined with a heap, this structure
  supports the shared intermediate results required by lazy
  evaluation.  Here, we take an alternative approach, compiling
  call-by-need lambda calculus directly to x64 assembly code.  By
  using shared environments to ensure bound terms are evaluated only
  once (a requirement of lazy evaluation), we achieve an
  implementation that is much simpler than any implementation of the
  call-by-need lambda calculus of which we are aware.  The semantics
  of the shared environment are defined by a new abstract machine, the Cactus
Environment Machine (CEM).  Beyond simplicity, the CEM approach
may lead to implementation efficiencies by exploiting tradeoffs
between flat and hierarchical storage structures.  [TODO: do we
believe this last sentence?]

%The compiler is remarkable primarily for its simplicity,
%which it achieves by implementing 
\end{abstract}

\section{Introduction} \label{sec:intro}

Lazy functional programming languages implement call-by-need lambda calculus
\cite{ariola1995call}. In these implementations, environments are
\emph{flat}, binding free variables to locations in the heap and
enforcing proper sharing.  
Here, we propose an alternative approach, which
implements the
sharing required for call-by-need semantics with a single
% large 
shared environment, which serves both as the environment for every
closure and as the heap. We refer to this structure as a \emph{cactus
  environment} and the resulting abstract machine as the \emph{cactus
  environment} ($\mathcal{CE}$) machine.

In this pearl, we focus only on the issue of implementation
simplicity, leaving careful
performance measurements for future work, although we believe that 
our approach could also lead to efficiency improvements.
%We do not attempt to show that the $\mathcal{CE}$ is better than existing
%approaches by traditional standards, such as speed, memory use,
%etc. Instead, we
Using $\mathcal{CE}$, we present a native code compiler that is so
simple that its essentail components fit on a single page, as shown in
Figure~\ref{fig:impl}.\footnote{For brevity we omit well-understood components---the
  parser, the translation to deBruijn indices, and the x64 preamble;
  the full compiler is available at \url{http://github.com/stelleg/cem_pearl}.}
The \texttt{compile} function takes as input a lambda term and
produces as output a list
of x64 basic blocks. Since we use only a small subset of the available x64
instructions, a simple string representation suffices.
\begin{code}
compile :: Expr -> [BasicBlock]
type BasicBlock = (Label, [Instr])
type Label = String
type Instr = String
\end{code}


\section{Preliminaries}

We assume familiarity with lambda calculus, environments, and closures.
We adopt standard Barendregt syntax [TODO add citation] for lambda terms,
$$ e ::=  \lambda x . e \; | \; e \;  e \; | \;
x $$, applications are left
associative, and lambda bodies extend as far as possible. 

Internally the compiler operates on lambda terms with deBruijn indices,
so we add the synatx $$ t ::= \lambda t \; | \; t \;
t \; | \; i $$ where $i \in \mathbb{N}$ replaces the variable name with an index
into the environment.   In the implementation,
\begin{code}
type Var = Int
data Expr = Lam Expr
           | Var Var
           | App Expr Expr
\end{code}
Instead of
binding named variables, here a variable indexes the depth of its binding lambda with
a natural number. For example $\lambda t.\lambda f.t$ in standard notation
becomes $\lambda\lambda1$ in lambda calculus with deBruijn indices. 
Examples are easier to write, read, and understand in standard lambda
calculus with named variables, and we will do so.

In mechanical evaluation of expressions, it would be too inefficient to perform
explicit substitution. To solve this, the standard approach uses
closures~\cite{landin1964mechanical, curien1991abstract, jonesstg, biernacka2007concrete}.
Closures combine an expression with an environment, binding the free
variables in the expression to other closures.  [TODO: Do we do this
in our implementation?]

\section{Environment Representation} \label{sec:env}

There are two common approaches that span the design space for environment
representation: \emph{flat} and \emph{shared} (also
known as linked)~\cite{appel1988optimizing, shao1994space}. In a flat
environment, each closure has its own record of what closures all
of its free variables are bound to. In a shared environment, parts
of that record are shared among multiple environments~\cite{appel1988optimizing,
shao1994space}. For example, consider the following closed term: $$(\lambda
x.(\lambda y.t) (\lambda z.t_1)) t_2$$ Assuming the term $t$ has both $x$ and
$y$ as free variables, we must evaluate it in the environment binding both $x$
and $y$.  Similarly, assuming $t'$ contains both $z$ and $x$ as free variables,
we must evaluate it in an environment containing bindings for both $x$ and $z$.
Thus, we can represent the closures for $t$ and $t_1$ as
$$t[x=t_2[\bullet], y=c]$$ and $$t_1[x=t_2[\bullet], z=c_1]$$ respectively. In a
flat representation, space is allocated separately for each
environment, and each closure contains
an explicit record of all of its free variables. Because of the nested scope of
the given term, $x$ is bound to the identical closure in each
environment and can also be represented by a shared, linked
structure, shown in the following
diagram:

\begin{center}
\begin{tikzpicture}[ 
  edge from parent path={(\tikzchildnode\tikzchildanchor) edge [-latex] (\tikzparentnode\tikzparentanchor)},
  level distance=1cm
]
\node (d) {$\bullet$} child{node (a) {$x=t_2[\bullet]$} child{node (b) {$y=c$}} child{node (c)
{$z=c_1$}}};

\end{tikzpicture}
\end{center}
In this scheme, each environment is represented by a linked list, with
the binding of $x$ shared between them. This simple example
illustrates a shared, linked environment ~\cite{appel1988optimizing}
and dates back to Landin's SECD machine~\cite{landin1964mechanical}.

The strengths and weaknesses of each approach are well known.  With a flat
environment variable lookup is implemented by a simple offset, or even
bound to a particular register~\cite{jonesstg, appel2006compiling},
but significant duplication can occur.  In a shared environment, unnecessary
duplication is avoided, but a cost is incurred on dereference whenever the
link structure must be traversed.

\subsection{Call-By-Need and the Heap}

%Because of the benefits of flat environments we listed above, and perhaps others we
%have overlooked, they 
Historically, flat environments have dominated call-by-need implementations
\cite{jonesstg, TIM, johnsson1984efficient,
  boquist1997grin}.\footnote{Possibly because
call-by-need evaluation was historically implemented with combinators, which take all their free variables
as formal parameters, and therefore suit flat environments well.}
Note that
for call-by-need languages, one must be careful when using flat
environments.  Because results from bound computations have to be shared across
closures referencing them, an implementation of call-by-need cannot copy closures
across environments. To illustrate this, consider again the earlier example.
If we copied closures, results from evaluation of $t_2[\bullet]$ from
dereferencing $x$ in $t$ would not be shared with the instances of $x$ in $t_1$.
This is fine for call-by-name, but to implement call by need, we must ensure
sharing.

The standard approach is to add a heap, which maps addresses to closures
\cite{jonesstg, TIM, johnsson1984efficient, sestoft} and then
modify the environments to map variables to addresses in the heap. The final
step adds update markers pointing to heap addresses on the stack, to update
the closures at those locations in the heap with their value once they have been
evaluated.  Reconsidering
%The reader may see where we are headed: 
the shared
environment diagram from the preceding section, if we updated the
$x=t_2[\bullet]$ with $t_2[\bullet]$'s corresponding value after it was
entered, the result would be shared between $t$ and $t_2$ without
an additional heap. This is what we now describe and implement.

\section{Shared Environment Call-By-Need} \label{sec:calc}

Here we describe the key idea of our approach, showing how
the shared environment described earlier can be applied to call-by-need
evaluation. We start with a well-known abstract machine for evaluating
call-by-name, the Krivine machine $K$, Figure~\ref{fig:Krivine}.
The mechanics are wonderfully simple: applications push their argument
onto the stack, abstractions pull an argument off the stack and move it into the
environment, and variables enter the closure at their index into the
environment.

\input{figures/figure-K}

Note that we have not specified how the environment is represented in the $K$
machine. To move towards the goal of sharing via environment structure,
we force the environment to be shared by creating a heap of environment
cells, and then extending each cell manually. This produces the $K'$ machine,
Figure~\ref{fig:K'}.

\input{figures/figure-Kprime}

Now, we have a heap of environments, which ensures that the
environments are shared when extended.  The $K'$ machine is identical to the $K$
machine, except that it forces this sharing of environments. 

[TO DO: SHOULD REPHRASE THE FOLLOWING SENTENCE--"Given our ... environments"... is awkward; "given" is a vague word]
Given our specified explicitly shared environments, we can add the necessary
machinery to implement full call-by-need semantics.
[CHECK THIS REWRITE:] To do so, we modify the $K'$ machine to make it lazy
in much the same way as Sestoft modified the Krivine machine~\cite{sestoft}.
%To do this, we add similar modifications to those others have added to the Krivine machine to make it lazy \cite{sestoft}.
We add update markers so that when a closure is entered, the
Var1 rule pushes an update marker to the corresponding location onto the stack.
Then, when the corresponding value is our current closure, it will pop the
update marker, and replace the closure on the heap with itself. This
call-by-need $L$ machine is specified in Figure~\ref{fig:L}.

\input{figures/figure-CE}

[TO DO: CONSISTENCY! IS IT $L$ or $\mathcal{CE}$? I vote for $L$]

We see from the semantics that all that has changed between the $K'$ and the
$\mathcal{CE}$ machine is the addition of the update marker to the Context
(which can be implemented as a normal stack), and the corresponding Update and
Var1 rules. This is in contrast to flat environment machines, such as the TIM~\cite{TIM}, which require more
complex machinery to ensure proper sharing.

This [TO DO: WHAT?] is similar to existing lazy variants of the Krivine machine, such as that
defined by Sestoft \cite{sestoft}, except that instead of leaving the
environment unspecified, and adding a heap of closures, we have forced the
environment to be shared, and used that shared environment to implement the
necessary sharing of results.\footnote{In fact, in the code comments of his
dissertation, Sestoft mentions possible environment sharing, but it is never
formalized.} 

\section{An Example}

We now turn to 
visualizations of example executions to provide intuition for how the $L$
machine operates. Figure~\ref{fig:states} shows a portion
of the execution trace during the evaluation of $(\lambda a.(\lambda b.b \; a)
(\lambda c.c \; a)) ((\lambda i.i) (\lambda j.j))$. 

\input{figures/figure-states}

We can see each rule in effect in our visualization, and how it changes the
state of the machine, directly. Hopefully this gives the reader some intiution
for how the machine operates, and how simple its rules are from a mechanical
point of view. [THIS IS NOT ENOUGH. You can't expect the reader to follow the diagrams
without any guidance. Include a narrative explanation of the whole figure here. You have enough
space for the figure caption on this page.]

\section{Correspondence to Call-By-Need Lambda Calculus}
To convince ourselves that the call-by-need
semantics are correctly implemented, we turn to the work of
Ariola et al.\cite{ariola1995call}, which specified an operational semantics for
call-by-need lambda calculus, Figure~\ref{fig:cbn}.

\input{figures/figure-cbn}

Case-wise induction on $\rightarrow_{N}$ gives an ordering to the closures
in the cactus environment. This is similar to the ordering from Lemma 8.1 of
Ref.~\cite{ariola1995call}.
Using this ordering, we define the function $flatten$, which takes any $(c, \mu)$
and flattens it accordingly. It does so by mapping any $(c, \mu)$ to a $\langle
\Phi \rangle t$, with the necessary variable hygiene. With this function, along
with structural induction on term size, it is not hard to show that
$\rightarrow_{\mathcal{CE}}$ bisimulates $\downarrow$.

%It is worth noting the similarity between these semantics for pedagogical
%purposes. 
Ariola et al.'s semantics throws out environment structure because
it doesn't need to keep it. Similarly the abstract machines derived from the
semantics (e.g., \cite{garcia2009lazy}) are forced to search for variables in the
environment---deBruijn indices are not an option. In contrast, our
machine retains that
structure in the cactus environment, which allows the use of deBruijn
indices and a straighforward low-level
implementation (next section).

\section{Implementation}\label{sec:impl}

% we already said this in the intro.
% To those readers who enjoy thinking in terms of machine code, it is
% already clear that the $\mathcal{CE}$ machine lends itself
% straightforwardly to implementation. Indeed, our
% entire native code compiler fits in a single page. 

First, some preliminaries. We target the x64 GNU assembler running on
Linux. Any x64 linux machine that has GNU ld should be able to run the
compiled examples.  We include in the data type a label for each basic
block so it can be jumped to directly, and we annotate each x64
instruction line with a comment summarizing its purpose.  We also
explain the simple mapping between each basic block and the
corresponding rule from the abstract machine specification.

The compile function takes a deBruijn indexed lambda term and
generates a list of basic blocks of x64 assembly code. In the case of the Var1
and Var2 rules, they are combined into a single basic block, which traverses the
environment linked list until it reaches a closure, then pushes an update marker
to that location. The App rule is the simplest: it consists of two instructions
that push the location of the argument term along with the current environment
pointer onto the stack. The Lam and Update rules are a little more complicated.
Recall that there is no rule in which we have a lambda as our current term, and
an empty stack. That is so, because that is the condition for termination. Thus, we
must add a check to see if the stack is empty (\texttt{checkTermBB}), and terminate if it
is (\texttt{termBB}). If it is not, then we check (\texttt{checkUpdateBB})
to see if it is an update marker (which
we have designated to be the \texttt{NULL} pointer or 0, as there cannot be a valid code
pointer to that location), and if it is,
update the marker and continue without changing the current closure. If it is
not an update marker, it is an argument closure, so we pop to extend our current
environment with it. 

\input{figures/figure-impl}

\section{Discussion}
[TODO: Deal with the ???]
The abstract machine as presented has some surprisingly nice properties and some
weaknesses. There is a sense in which the CEM ???? is \emph{lazier} than flat
environment implementations. Instead of spending time allocating a flat
environment so that if a closure is entered multiple times ???? variable lookup, our
implementation does the minimal amount of preparation work. Of course, this
comes with the previously mentioned cost of non-constant variable lookup time. 

Instead????, we hope that this simple implementation could be of 
pedagogical value. Furthermore, we believe it could make for a good target
for abstraction thanks to its simplicity, following the approach of Van Horn et al.
\cite{van2010abstracting}. [????What does it mean to be a target for abstraction? Explain.]

\section{Conclusion and Future Work}
We hope that the reader is convinced of the elegance of this approach to implementing
call-by-need, and that this use of shared environments can lead to
improved future implementations. We are not aware of any call-by-need
implementations that are as simple as the one described here.
%this simple, and so end with a quote from Walt Whitman: "Simplicity is the glory
%of expression".
Beyond simplicity and elegance, however, 
we are actively exploring compile-time optimizations to make this abstract machine
practically usable for lazy language implementation. [ANYTHING ELSE?]

% We recommend abbrvnat bibliography style.
\bibliographystyle{abbrvnat}
\bibliography{annotated}

\end{document}

